---
title: "dplyr exercises"
author: "Stefan"
format: html
editor: visual
---

## Setup

```{r}
library("tidyverse")


# load data
adsl <- haven::read_sas("data/adsl.sas7bdat") %>% 
  mutate_if(is_character, na_if, "")
```

## Data wrangling with dplyr

Load the `adsl` data-frame and select the following variables:

-   USUBJID

-   ARM

-   SEX

-   AGE

-   AGEU

-   AGEGR1

-   COUNTRY

-   EOSSTT

```{r}
#| code-fold: true
#| code-summary: "Show solution"
adsl %>% 
  select(USUBJID, ARM, SEX, starts_with("AGE"), COUNTRY, EOSSTT)

```

On the selected variables, include only patients in the placebo arm who are 66, 77, 88, or 99 years old.

```{r}
#| code-fold: true
#| code-summary: "Show solution"
adsl %>% 
  select(USUBJID, SEX, ARM, EOSSTT, starts_with("AGE")) %>% 
  filter(ARM == "Placebo",
         AGE %in% c(6:9*11))
```

Further include the variable TRTSDTM (datetime of first exposure to treatment) and sort the previous data-frame according to this variable from most recent to least recent first exposure.

```{r}
#| code-fold: true
#| code-summary: "Show solution"
adsl %>% 
  select(USUBJID, SEX, ARM, EOSSTT, starts_with("AGE"), TRTSDTM) %>% 
  filter(ARM == "Placebo",
         AGE %in% c(6:9*11)) %>% 
  arrange(desc(TRTSDTM))
```
