---
title: "mutate exercises"
author: "Thomas Neitmann"
format: html
editor: visual
---

## Setup

```{r message=FALSE}
library(dplyr)
library(lubridate)
dm <- readRDS("data/dm.rds")
ae <- readRDS("data/ae.rds")
```

## Exercise 1

A treatment emergent adverse event is defined as an adverse event whose start date is on or after the treatment start date (`TRTSDT`) and at the latest starts 7 days after the treatment end date (`TRTEDT`). Given this definition calculate `TRTEMFL`.

*Hint*: Turn the `--DTC` variables into proper dates first using the `ymd()` function.

```{r}
#| code-fold: true
#| code-summary: "Show solution"
ae %>% 
  mutate(
    ASTDT = ymd(AESTDTC),
    AENDT = ymd(AEENDTC),
    TRTEMFL = if_else(ASTDT >= TRTSDT & ASTDT <= TRTEDT + 7, "Y", NA_character_)
  ) %>% 
  select(USUBJID, ASTDT, AENDT, TRTSDT, TRTEDT, TRTEMFL)
```

## Exercise 2

Create a new variable `REGION1` based upon `COUTRY` as shown in the table below.

| Countries                           | Region        |
|-------------------------------------|---------------|
| Mexico, USA, Canada                 | North America |
| Spain, Greece, Germany, Switzerland | Europe        |
| China, Japan                        | Asia          |

```{r}
#| code-fold: true
#| code-summary: "Show solution"
dm %>% 
  mutate(
    REGION1 = case_when(
      COUNTRY %in% c("Mexico", "USA", "Canada") ~ "North America",
      COUNTRY %in% c("Spain", "Greece", "Germany", "Switzerland") ~ "Europe",
      COUNTRY %in% c("China", "Japan") ~ "Asia"
    )
  ) %>% 
  select(USUBJID, COUNTRY, REGION1)
```
