---
title: "tidyr exercises"
format: html
editor: visual
---

## Setup

```{r, warning = FALSE, message = FALSE}
library(tidyverse)
library(admiral)
library(admiral.test)
library(dplyr)
library(tidyr)

# load data
ex <- admiral_ex
dm <- admiral_dm
ds <- admiral_ds
suppds <- admiral_suppds
```

## Pivoting with tidyr

Load the `ex` data-frame from `admiral_ex`and select the following variables:

-   USUBJID

-   EXTRT

-   VISIT

-   EXSTDTC

```{r}
#| code-fold: true
#| code-summary: "Show solution"
ex %>% 
  select(USUBJID, EXTRT, VISIT, EXSTDTC)
```

Using `pivot_wider()` create a table that would shaped this way

| USUBJID | EXTRT | BASELINE | WEEK 2 | WEEK 24 |
|---------|-------|----------|--------|---------|
| ...     | ...   | ...      | ...    | ...     |

```{r}
#| code-fold: true
#| code-summary: "Show solution"
ex %>% 
  select(USUBJID, EXTRT, VISIT, EXSTDTC) %>% 
  pivot_wider(names_from = "VISIT", values_from = "EXSTDTC")

```

Load the `dm` data-frame from `admiral_dm`and select the following variables:

-   USUBJID

-   RACE

-   SEX

```{r}
#| code-fold: true
#| code-summary: "Show solution"
dm %>% 
  select(USUBJID, RACE, SEX)
```

Using `pivot_longer()` create a table that would shaped this way

| USUBJID | VAR  | VAL   |
|---------|------|-------|
| 1001    | RACE | WHITE |
| 1001    | SEX  | M     |

```{r}
#| code-fold: true
#| code-summary: "Show solution"
dm %>% 
  select(USUBJID, RACE, SEX) %>% 
  pivot_longer(cols = c(RACE, SEX),
               names_to = "VAR",
               values_to = "VAL")

```

## Joining using dplyr

Load the `ds` data-frame from `admiral_ds` and `suppds` data-frame from `admiral_suppds`. Prior to joining the two datasets together, we may need to do some cleaning of the data on `suppds`.

-   Filter IDVAR for "DSSEQ"

-   Mutate IDVARVAL from type character to type numeric.

-   Select USUBJID IDVARVAL QNAM QLABEL QVAL

```{r}
#| code-fold: true
#| code-summary: "Show solution"

suppds <- suppds %>% 
  filter(IDVAR == "DSSEQ") %>% 
  mutate(IDVARVAL = as.numeric(IDVARVAL)) %>% 
  select(USUBJID, IDVARVAL, QNAM, QLABEL, QVAL)

suppds
```

Join the two tables together using `USUBJID` and `DSSEQ` as the key joining variables.

```{r}
#| code-fold: true
#| code-summary: "Show solution"
ds %>% 
  left_join(suppds, by = c("USUBJID" = "USUBJID", "DSSEQ" = "IDVARVAL"))
```
